---
title: \vspace{-1.5cm} Wilderness Air Quality Values \break for `r input$x`
date: 'Report prepared: `r format(Sys.time(), "%d %B %Y")`'
toccolor: 'black'
header-includes:
   - \usepackage{hyperref} \hypersetup{pdfstartview={XYZ null null 1.00}, colorlinks = true, linkcolor = red, urlcolor = red}
   - \usepackage{float} \floatplacement{figure}{H}
   - \usepackage[font=small,skip=0pt]{caption}
   - \usepackage{wrapfig}
   - \usepackage{booktabs}
   - \usepackage{longtable}
---

```{r setup-options, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '')
has_ttest <- FALSE
```

```{r function-setup-functions, include=FALSE}
### set graphical parameters
`set_par` <- function (panels = 1, pty = 's', ...) {
    mgp <- c(2, 0.4, 0)
    mar <- c(4, 4, 0.5, 0.5)
    oma <- c(0, 0, 0, 0)
    auto_rowcol <- function(n = panels) {
        if (n <= 3) 
            c(1, n)
        else if (n <= 6) 
            c(2, (n + 1)%/%2)
        else if (n <= 12) 
            c(3, (n + 2)%/%3)
        else c(ceiling(n/(nr <- ceiling(sqrt(n)))), nr)
    }
    mfrow <- auto_rowcol()
    if (panels > 4) 
        panels <- 4
    switch(as.character(panels), `4` = par(mfrow = mfrow, 
        mgp = mgp, mar = mar, pty = pty, oma = oma, bty = 'L', 
        las = 1, cex.lab = 1.2, tcl = -0.2, ...), `3` = par(mfrow = mfrow, 
        mgp = mgp, mar = mar, pty = pty, oma = oma, bty = 'L', 
        las = 1, cex.lab = 1.4, cex.axis = 1.2, tcl = -0.2, ...), 
        `2` = par(mfrow = mfrow, mgp = mgp, mar = mar, 
            pty = pty, oma = oma, bty = 'L', las = 1, cex.axis = 0.85, 
            tcl = -0.2, ...), `1` = par(mfrow = mfrow, 
            mgp = mgp, mar = mar, pty = pty, oma = oma, bty = 'L', 
            las = 1, cex.axis = 0.85, tcl = -0.2, ...))
}
### get years of sampling
`get_years` <- function (pick, ...) {
        rng <- range(d[d$wa==pick, 'year'], na.rm=TRUE)  ### <<--------------------
        if (any(!is.finite(rng))) c(NA,NA) else rng
}
### get recent decade
`get_recent_decade` <- function (pick, ...) {
     s <- sort(unique(a[a$warea==pick, 'tenyr']))
     s <- s[length(s)]
     if (any(!is.finite(s))) c(NA) else s
}
### get number of plots
`get_n` <- function (pick, ...) {
        n <- unique(a[a$warea==pick, 'n_plots'])
        if (any(!is.finite(n))) c(NA) else n
}
### table function with 5-y mean
`get_tab` <- function (pick, xvar, col.names, caption, format){
        tab <- a[which(a$warea==pick), xvar]
        if (missing(col.names)) col.names <- xvar
        if (missing(caption))   caption   <- NULL
        if (missing(format))    format    <- 'pandoc'
        j   <- xvar[xvar != 'tenyr'] # dont consider year column
        tab <- tab[rowSums(is.na(tab[,j])) != ncol(tab[,j]),]
        tab[tab %in% c('NA','***NA***','NaN','***NaN***')] <- NA 
        knitr::kable(x         = tab,
                     format    = format,
                     row.names = F,
                     align = c('l', rep('r',NCOL(tab)-1)),
                     longtable = T,
                     booktabs  = T,
                     caption   = caption,
                     col.names = col.names)
}
```

\vspace*{\fill}
\begin{wrapfigure}{l}{1.8cm}
\includegraphics{usfs.png}
\end{wrapfigure}

\  

Air Resource Management Program  
USDA Forest Service  
1400 Independence Ave SW  
Washington, DC 20250  

\  

\newpage

<!-- # Overview -->

<!-- __*Evaluating status and trends in air quality and terrestrial condition using lichen WAQVs*__ -->

<!-- This report uses the __*Lichen Wilderness Air Quality Value (WAQV)*__ to indicate status and trends in terrestrial ecosystem condition in `r input$x`.  This WAQV meets Wilderness Stewardship Performance reporting requirements for the Air Element.  -->

<!-- Fertilizing and acidifying air pollutants alter soil and water chemistry, which can change the diversity and composition of microbes, invertebrates, vegetation and wildlife. Together with ozone and air toxics (not reported here) they are the air pollutants most likely to harm terrestrial and aquatic wilderness ecosystems.  -->

<!-- Because most harm occurs as pollutants are deposited to the environment, air pollution levels and ecosystem responses typically use units of deposition, i.e., kilograms of __*total nitrogen*__ (fertilizing pollution) or __*total sulfur*__ (acidifying pollution) per hectare per year.  The short way to denote these units is kg N ha^-1^ y^-1^ or kg S ha^-1^ y^-1^. -->

<!-- Ecosystem responses to air pollution can be quantified by single-species measures (e.g., growth or mortality), or by community measures (e.g., richness or community composition).  -->

<!-- A __*critical load (CL)*__ is the amount of pollutant deposition below which no ecosystem harm occurs. CLs exist for soils, trees, herbs, and lichens. As air pollution surpasses a given CL, both the amount of __*exceedance*__ (kg ha^-1^ y^-1^) and the __*risk*__ of anticipated harm to ecosystems increase. As air quality improves, we expect reductions in CL exceedances and improvements in ecosystem responses. Thus, continued monitoring is a key tool for quantifying how air quality relates to ecosystem improvement, maintenance, or decline. -->

<!-- Epiphytic macrolichens (large lichens growing on trees and shrubs) are a preferred indicator of air pollution. USFS Forest Inventory \& Analysis protocols have been used to survey lichens for air quality and climate status and trends across most of the forested United States. Because lichens rely on air for their nutrition and lack specialized tissues to exclude pollution, the number and type of species present reflects air quality conditions. Unlike the buffering effect of soils which protects rooted plants, lichens respond directly to the amount of pollution in the air. Thus lichen CLs are the proverbial canary in the coal mine. -->

<!-- Lichens are also an integral part of forested ecosystems. They are composite organisms of a fungus and a green alga or a cyanobacterium or both. The fungus provides nutrients from the air and a protective environment, the algae photosynthesize, and the cyanobacteria fix nitrogen from the air and will also photosynthesize if no green alga is present. North America possesses >5400 species. Cyanobacterial lichens (__*cyanolichens*__) are highly preferred by invertebrates for their moderated moisture and higher nutrient content compared to green algal lichens, and cyanolichens contribute hugely to forest nutrient cycling. __*Forage lichens*__ are beard-like, shrubby species prized by ungulates, rodents, and birds for food and nesting materials. __*Matrix lichens*__ are all other leaf-like and shrubby species primarily used by invertebrates as habitat and birds for nesting. -->

<!-- This report summarizes status and trends in air quality and its effects in 10-year increments including: -->
<!-- * atmospheric deposition of nitrogen and of sulfur (CMAQ v5.02, lichen community air scores) -->
<!-- * exceedances of lichen N and S critical loads -->
<!-- * risks to lichen biodiversity, sensitive lichens, forage lichens, cyanolichens, and individual species.  -->
<!-- Status is provided as a concern level: none, minimal, moderate, or high; trends as: no change, increasing, or decreasing.  -->
<!-- See **[Further information](#further-information)** section for additional documentation of these methods. -->


<!-- \newpage -->


# Summary for `r input$x`

```{r years_and_plot_count}
rng <- get_years(pick = rx())
n   <- get_n(pick = rx())
dec <- get_recent_decade(pick = rx())
```

Number of surveyed plots:   **`r n`**  
Field sampling dates: **`r rng[1]` to `r rng[2]`**  
Most recent ten-year sampling period: **`r dec`**  

## Nitrogen snapshot

```{r plot-plot_valuebar_N, warning=FALSE, message=FALSE, fig.height=2.25, fig.width=6.0, fig.cap='Observed N deposition (arrow) for the most recent 10-year sample period, in relation to levels of concern.  Levels of concern (Acceptable, Marginal, Concerning, Poor, Very poor) are defined by the expected decline of sensitive lichens (0, 0-20, 20-50, 50-80, and >80 percent declines established for critical loads).'}
### plot the valuebar for observed CMAQ deposition
`plot_valuebar` <- function(pick, brk=b, select='n', ...){
     b      <- brk
     x      <- a[which(a$warea==pick),
                 paste0('cmaq_',select,'_3yroll.mean')]
     arrlab <- 'OBSERVED\nCMAQ deposition'
     if(all(is.na(x))) { # try lichen airscores if CMAQ is missing
          x      <- a[which(a$warea==pick),
                      paste0(select,'_airscore.mean')]
          arrlab <- 'OBSERVED\nlichen airscores'
     } 
     if(all(is.na(x))) { # no more tries
          x <- 999
     }
     obs  <- x[length(x)] # take most recent one
     par(mar=c(4,0.3,0.3,0.3), las=1, tcl=-0.2, mgp=c(1.8,0.4,0))
     n    <- 100
     xseq <- seq(-0.5, 20.0, len=n)
     plot(xseq, seq(1, 10, len=n), type='n', bty='n',
          yaxt='n', ylab='', ylim=c(1,13.4), yaxs='i', xaxs='r', ...)
     cc   <- viridis::inferno(5,begin=0.1,end=0.9,alpha=seq(0.7,1,len=5))
     rect(-9,   -10, b[1], 20, col=cc[1], border=NA)
     rect(b[1], -10, b[2], 20, col=cc[2], border=NA)
     rect(b[2], -10, b[3], 20, col=cc[3], border=NA)
     rect(b[3], -10, b[4], 20, col=cc[4], border=NA)
     rect(b[4], -10, 99,   20, col=cc[5], border=NA)
     text(x=b[1] - 0.5*(b[1]-(-1)), y=13.1, 'Acceptable', pos=1, cex=0.6)
     text(x=b[2] - 0.5*(b[2]-b[1]), y=13.1, 'Marginal', pos=1, cex=0.6)
     text(x=b[3] - 0.5*(b[3]-b[2]), y=13.1, 'Concerning', pos=1, cex=0.6)
     text(x=b[4] - 0.5*(b[4]-b[3]), y=13.1, 'Poor', pos=1, cex=0.6)
     text(x=21   - 0.5*(21-b[4]),   y=13.1, 'Very poor', pos=1, cex=0.6)
     if (obs >= 999) {
          text(x=10, y=10, 'No data available...', pos=1, cex=0.7, font=2)
     } else if (obs > 20 & obs < 999) {
          arrows(20.2, 6.5, 20.2, 1, length=0.20, angle=15, lwd=3, col=1)
          text(x=18.8, y=10, arrlab, pos=1, cex=0.7, font=2)
     } else {
          arrows(obs, 6.5, obs, 1, length=0.20, angle=15, lwd=3, col=1)
          text(x=obs, y=10, arrlab, pos=1, cex=0.7, font=2)
     }
}
b <- c(2.5, 6, 9.9, 14) # loosely based on 20/50/80
plot_valuebar(pick = rx(), brk=b, select='n',
              xlab=expression(N~deposition~'('*kg~ha^-1~yr^-1*')'))
```

\  

## Sulfur snapshot

```{r plot_valuebar_S, warning=FALSE, message=FALSE, fig.height=2.25, fig.width=6.0, fig.cap='Observed S deposition for the most recent 10-year sample period, in relation to levels of concern.  Levels of concern (Acceptable, Marginal, Concerning, Poor, Very poor) are defined by the expected decline of sensitive lichens (0, 0-20, 20-50, 50-80, and >80 percent declines established for critical loads).'}
### plot the valuebar for observed CMAQ S deposition
b <- b+(b*0.05) # new (loosely based on 20/50/80)
plot_valuebar(pick = rx(), brk=b, select='s',
              xlab=expression(S~deposition~'('*kg~ha^-1~yr^-1*')'))
```


\newpage


```{r check_ttest, include=FALSE}
x   <- d[d$wa==rx(), ]
u   <- sort(unique(na.omit(x[,'tenyr']))) # get unique time periods
len <- length(u)
if(len < 2) {
  has_ttest <- FALSE
} else {
  u <- u[(len-1):(len)]   # take *most recent* two periods
  x <- droplevels(x[x$tenyr %in% u,,drop=T])
  x <- x[order(x$tenyr), ]
  ### either tenyr have *ALL* NA? (ttest still works if *SOME* NA)
  `f` <- function(k) all(is.na(k))
  ### control logic to select CMAQ > AIRSCORE > NONE (nitrogen)
  if (!any(tapply(x$cmaq_n_3yroll, list(x$tenyr), f))) {
    fmla_n <- as.formula(cmaq_n_3yroll ~ tenyr)
  } else if (!any(tapply(x$n_airscore, list(x$tenyr), f))) {
    fmla_n <- as.formula(n_airscore ~ tenyr)
  } else {
    fmla_n <- NULL
  }
  ### control logic to select CMAQ > AIRSCORE > NONE (sulfur)
  if (!any(tapply(x$cmaq_s_3yroll, list(x$tenyr), f))) {
    fmla_s <- as.formula(cmaq_s_3yroll ~ tenyr)
  } else if (!any(tapply(x$s_airscore, list(x$tenyr), f))) {
    fmla_s <- as.formula(s_airscore ~ tenyr)
  } else {
    fmla_s <- NULL
  }
  ### ensure minimum sample size
  `get_n_per_grp` <- function (x, na.rm = TRUE) {
    n  <- length(x) - sum(is.na(x))
    return(c(n=n))
  }
  n_n <- tryCatch(aggregate(fmla_n, data=x, get_n_per_grp),
                  error=function(e) return(NULL))
  n_s <- tryCatch(aggregate(fmla_s, data=x, get_n_per_grp),
                  error=function(e) return(NULL))
  ### control logic, if both are NULL or too-small sample-size, 
  ###     then dont do the ttests
  if ((is.null(fmla_n) & is.null(fmla_s)) | 
      (any(n_n[,2] < 2) & any(n_s[,2] < 2))) {
    has_ttest <- FALSE
  } else {
    ### UNPAIRED 2-sample ttest (plots NOT paired...)
    t1 <- tryCatch(t.test(fmla_n,data=x),
                   error=function(e)return(NULL))
    t3 <- tryCatch(t.test(fmla_s,data=x),
                   error=function(e)return(NULL))
    ### descriptive summary table
    `f` <- function (x, na.rm = TRUE) {
      if (!is.numeric(x)) return(NULL)
      m  <- mean(x, na.rm = na.rm)
      s  <- stats::sd(x, na.rm = na.rm)
      na <- sum(is.na(x))
      n  <- length(x) - na
      se <- s/sqrt(n - 1)
      ci <- 1.96 * se
      return(c(mean=round(m,3), ci=round(ci,3), n=n))
    }
    tab1 <- tryCatch(do.call(data.frame, 
                             aggregate(fmla_n, data=x, f)),
                     error=function(e)return(NULL))
    tab3 <- tryCatch(do.call(data.frame, 
                             aggregate(fmla_s, data=x, f)),
                     error=function(e)return(NULL))
    has_ttest <- TRUE
    
    ### if ttests failed then cancel ttests
    if (is.null(t1) & is.null(t3)) {
      has_ttest <- FALSE
    }
  }
}
```

`r if(has_ttest) {paste0('## Significant trends\n\nFrom the decade ', u[1], ' to ', u[2], ', a two-sample _t_-test revealed ', ifelse(t1$p.value < 0.05,'a ','NO '), 'SIGNIFICANT trend in N deposition, and ', ifelse(t3$p.value < 0.05,'a ','NO '), 'SIGNIFICANT trend in S deposition. ', 'See figure for test results.')}`

```{r plot_ttest, eval=has_ttest, echo=FALSE, fig.width=6.5, fig.height=3.5, fig.cap='Results of _t_-tests evaluating air-quality trends over time.'}
### PLOT
`plot_ci` <- function (tab, mod, ylab, ...) {
     nm   <- tab[,1]
     y    <- tab[,2]
     yci  <- tab[,3]
     n    <- tab[,4]
     x    <- 1:2
     ylm  <- c(0, # min(c(y, y - yci), na.rm=T) - 0.4,
               max(c(y, y + yci), na.rm=T) + 0.1)
     if (all(is.infinite(ylm))) return(NULL)
     if (anyNA(ylm)) { ylm[which(is.na(ylm))] <- NULL }
     if (missing(ylab)) {
          ylb <- expression(Air~score~(kg~ha^'-1'~y^'-1'))
     } else {
          ylb <- ylab
     }
     plot(x=x, y=tab[,2], pch=16, ylab=ylb, xlab='', xaxt='n',
          xlim=c(0.4,2.6), ylim=ylm, ...)
     xlb <- paste0(nm,'\n','n = ', n)
     axis(1, at=x, labels=xlb, tick=F, line=1, cex.axis=0.9)
     w <- 0.05
     segments(x0=x,   x1=x,   y0=y-yci, y1=y+yci, lwd=1.75)
     segments(x0=x-w, x1=x+w, y0=y+yci, y1=y+yci, lwd=1.75)
     segments(x0=x-w, x1=x+w, y0=y-yci, y1=y-yci, lwd=1.75)
     `add_text` <- function (x, y, labels, ...) {
          text(x = graphics::grconvertX(x, from = "npc", to = "user"),
               y = graphics::grconvertY(y, from = "npc", to = "user"),
               labels = labels, adj = 0, ...)
     }
     add_text(0.05,0.19,
              lab=paste0('t = ', formatC(round(mod$statistic,1),digits=3)),
              cex=0.8)
     add_text(0.05,0.12,
              lab=ifelse(mod$p.value < 0.01, 'p < 0.01',
                         paste0('p = ',formatC(round(mod$p.value,2),digits=4))),
              cex=0.8)
     add_text(0.05,0.05,
              lab=paste0('Est. diff. = ',
                         formatC(round(diff(mod$estimate),1),digits=2)),
              cex=0.8)
}
### setup labeling
if (grepl('cmaq', names(tab1)[2])) {
     ylb_n <- expression(CMAQ~value~(kg~N~ha^'-1'~y^'-1'))
} else if (grepl('airscore', names(tab1)[2])) {
     ylb_n <- expression(N~air~score~(kg~N~ha^'-1'~y^'-1'))
} else {
     ylb_n <- NA
}
if (grepl('cmaq', names(tab3)[2])) {
     ylb_s <- expression(CMAQ~value~(kg~S~ha^'-1'~y^'-1'))
} else if (grepl('airscore', names(tab3)[2])) {
     ylb_s <- expression(S~air~score~(kg~S~ha^'-1'~y^'-1'))
} else {
     ylb_s <- NA
}
### plot
set_par(2)
par(mgp=c(1.3,0.4,0))
try(plot_ci(tab1, t1, ylab=ylb_n), silent=TRUE)
try(plot_ci(tab3, t3, ylab=ylb_s), silent=TRUE)
rm(get_n_per_grp, plot_ci, len, x, u, f, t1, t3, tab1, tab3, ylb_n, ylb_s, fmla_n, fmla_s)
```

`r if(has_ttest) {"Power analysis: a _t_-test would require _n_ = 17 plots per time interval to detect a change of 1 kg ha^-1^ y^-1^ with 80\\% probability.\\newpage"}`


## Area map

```{r plot-map, fig.height=6.5, fig.width=6.49, fig.cap='Local area map.  Red outline is Wilderness boundary, black points are survey plot locations, shaded contours are terrain elevations above sea level.  Boundaries are approximate.', warning=FALSE}
###  map wilderness boundary and sampling locations
`map_points` <- function (pick=pick, pts=d, shp=w, ras=dem, 
                          lwd.cont=1, ...) {
  aea_prj <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5
        +lon_0=-110 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m"
  trg_prj <- projection(shp)
  shp  <- shp[shp$NAME == pick,]  # select your wilderness
  pts  <- pts[pts$wa == pick,]    # select your Wilderness
  if(all(is.na(pts$lon)) | all(is.na(pts$lon))) {
    e <- extent(shp) * 1.25
    has_points <- FALSE
  } else {
    has_points <- TRUE
    `make_xy` <- function(xy, crs, ...){
      xy <- as.data.frame(xy)
      xy <- xy[!(is.na(xy$lon) | is.na(xy$lat)),] # rm NAs
      coordinates(xy) <- ~lon+lat
      proj4string(xy) <- CRS('+init=epsg:4269') # NAD83 dec degree
      return(spTransform(xy, crs))
    }
    ppts <- make_xy(pts, crs=trg_prj)
    near <- c(rgeos::gDistance(spTransform(ppts, CRS(aea_prj)),
                               spTransform(shp, CRS(aea_prj)),
                               byid=T) < 3000) # reject pts > 3 km
    ppts <- ppts[near,]
    e <- cbind(matrix(extent(shp)),matrix(extent(ppts)))
    e <- c(min(e[1,]),max(e[2,]),min(e[3,]),max(e[4,])) # combined ext
    `haversine` <- function(loc1, loc2) { # geographic distance
      lat1 <- loc1[1] * pi / 180
      lon1 <- loc1[2] * pi / 180
      lat2 <- loc2[1] * pi / 180
      lon2 <- loc2[2] * pi / 180
      R    <- 6371.0  # avg earth radius [km]
      dlat <- lat2 - lat1
      dlon <- lon2 - lon1
      a <- sin(dlat/2)^2 + cos(lat1) * cos(lat2) * sin(dlon/2)^2
      c <- 2 * atan2(sqrt(a), sqrt(1-a))
      return(R * c)
    }
    dy <- haversine(c(e[3],e[1]), c(e[4],e[1])) # kilometers y-axis
    dx <- haversine(c(e[3],e[1]), c(e[3],e[2])) # kilometers x-axis
    e  <- extent(e)
    if (dy > dx) {
      e[1:2] <- matrix(e * dy/dx)[1:2,] # expand x
    } else {
      e[3:4] <- matrix(e * dx/dy)[3:4,] # expand y
    }
    dx <- as.numeric( # how large a scale bar to make?
      as.character(
        cut(dx,breaks=c(0,5,50,200,Inf),labels=c(1,5,10,50),incl=T)
      )
    )
    e  <- extent(e) * 1.25 # slight expansion
  }
  ras <- disaggregate(crop(ras, e), 10, method='bilinear')
  ### custom function for filled contours....
  `fillcont` <- function (x, maxpixels = 1e+05, ...) {
    `fc` <-  function (
      x = seq(0, 1, length.out = nrow(z)),
      y = seq(0, 1, length.out = ncol(z)),
      z,
      xlim = range(x, finite = TRUE),
      ylim = range(y, finite = TRUE),
      zlim = range(z, finite = TRUE),
      levels = pretty(zlim, nlevels),
      nlevels = 20,
      color.palette = function(n) viridis::viridis(n,begin=0,end=0.9),
      col = color.palette(length(levels) - 1), ...) {
      if (missing(z)) {
        if (!missing(x)) {
          if (is.list(x)) {
            z <- x$z
            y <- x$y
            x <- x$x
          }
          else {
            z <- x
            x <- seq.int(0, 1, length.out = nrow(z))
          }
        }
        else stop("no 'z' matrix specified")
      }
      else if (is.list(x)) {
        y <- x$y
        x <- x$x
      }
      if (any(diff(x) <= 0) || any(diff(y) <= 0))
        stop("increasing 'x' and 'y' values expected")
      plot.new()
      plot.window(xlim, ylim, "", xaxs = "i", yaxs = "i", asp = NA)
      .filled.contour(x, y, z, levels, col)
    }
    x <- sampleRegular(x, maxpixels, asRaster = T, useGDAL = T)
    X <- xFromCol(x, 1:ncol(x))
    Y <- yFromRow(x, nrow(x):1)
    Z <- t(matrix(getValues(x), ncol = x@ncols, byrow=T)[nrow(x):1,])
    lonlat <- couldBeLonLat(x, warnings = FALSE)
    asp <- list(...)$asp
    if (is.null(asp)) {
      if (lonlat) {
        ym <- mean(c(x@extent@ymax, x@extent@ymin))
        asp <- 1/cos((ym * pi)/180)
      }
      else {
        asp <- 1
      }
      fc(x = X, y = Y, z = Z, asp = asp, ...)
    }
    else {
      fc(x = X, y = Y, z = Z, ...)
    }
  }
  fillcont(ras, nlevels=20, lwd=lwd.cont)
  plot(shp, col='transparent', border='red', add=T, lwd=2)
  if (has_points) {
    points(ppts, pch=16, col='#000000', cex=1)
    scalebar(d=dx, xy=c(e[1] + abs(diff(e[1:2]))*0.05,
                        e[3] + abs(diff(e[3:4]))*0.08),
             type='bar', divs=2, below=paste0('\n', dx, ' km'),
             adj=c(0.5,-1), lab='', lwd=2, cex=0.7, font=2)    
  }
}
### error catch -- dont plot if valid name not in shapefile
`try_map` <- function() {
     tryCatch(map_points(pick = rx(), pts=d, shp=w, ras=dem),
              error = function(e) {
                   plot(1:10, 1:10, type='n', axes=FALSE, ann=FALSE)
                   text(5,5,labels='Map not available',adj=c(0.5,0.5))
              })
}
try_map()
# map_points(pick = rx(), pts=d, shp=w, ras=dem)
```


\newpage


# Site scores

You can think of site scores (or "air scores") as the "average" position of species along an air quality gradient.  Technically, these are the community-weighted mean of species' sensitivities, where sensitivities are defined as the value of N or S deposition (kg ha^-1^ y^-1^) at which each species was most frequently detected.  We report both raw and climate-adjusted scores, and their nationwide ranking as a percentile.^[See **["Indicator: Air scores"](#indicator-air-scores)** section for more information.].

## Nitrogen

### N: mean deposition
```{r table_n_site}
get_tab(pick = rx(), 
        xvar=c('tenyr','n_airscore.mean','nadj.mean','cmaq_n_3yroll.mean'), 
        col.names=c('Ten-year interval', 'Lichen airscore (raw)', 'Lichen airscore (adj)', 'CMAQ model'), 
        caption= 'Site scores for estimated nitrogen deposition (N), as a *mean* (kg ha^-1^ y^-1^).')
```

### N: percentiles
```{r table_n_site_perc}
get_tab(pick = rx(), 
        xvar=c('tenyr','p_n_airscore.mean','p_n_adj.mean','p_n_cmaq.mean'), 
        col.names=c('Ten-year interval', 'Lichen airscore (raw)', 'Lichen airscore (adj)', 'CMAQ model'),
        caption= 'Site scores for estimated nitrogen deposition (N), as a *percentile* (0-100).')
```

## Sulfur

### S: mean deposition
```{r table_s_site}
get_tab(pick = rx(), 
        xvar=c('tenyr','s_airscore.mean','sadj.mean','cmaq_s_3yroll.mean'), 
        col.names=c('Ten-year interval', 'Lichen airscore (raw)', 'Lichen airscore (adj)', 'CMAQ model'), 
        caption= 'Site scores for estimated sulfur deposition (S), as a *mean* (kg ha^-1^ y^-1^).')
```

### S: percentiles
```{r table_s_site_perc}
get_tab(pick = rx(), 
        xvar=c('tenyr','p_s_airscore.mean','p_s_adj.mean','p_s_cmaq.mean'), 
        col.names=c('Ten-year interval', 'Lichen airscore (raw)', 'Lichen airscore (adj)', 'CMAQ model'),
        caption= 'Site scores for estimated sulfur deposition (S), as a *percentile* (0-100).')
```


\newpage


# Critical loads exceedances

Exceedances are the departure of raw or climate-adjusted lichen airscores from established lichen-based critical loads^[See **["Critical loads we used"](#critical-loads-we-used)** section.], expressed as the mean of a ten-year period.  Negative values indicate the "safety margin" below the critical load threshold, positive values indicate exceedance above the threshold. 

## Nitrogen

### N exceedance: raw lichen airscores

```{r table_exceedance_n_A}
get_tab(pick = rx(), 
        xvar=c('tenyr', 'n_raw_ex_sr.mean', 'n_raw_ex_sr_sens.mean',
               'n_raw_ex_sr_forag.mean', 'n_raw_ex_sr_cyano.mean', 
               'n_raw_ex_compn.mean'), 
        col.names=c('Ten-year interval', 'Richness', 'Sensitive richness', 
                    'Forage lichen abund', 'Cyanolichen abund', 
                    'Comm comp'), 
        caption= 'CL exceedances for mean nitrogen deposition (kg N ha^-1^ y^-1^), based on raw lichen airscores.')
```

### N exceedance: climate-adjusted lichen airscores

```{r table_exceedance_n_B}
get_tab(pick = rx(), 
        xvar=c('tenyr', 'n_adj_ex_sr.mean',       
               'n_adj_ex_sr_sens.mean',   
               'n_adj_ex_sr_forag.mean',  
               'n_adj_ex_sr_cyano.mean', 
               'n_adj_ex_compn.mean'),
        col.names=c('Ten-year interval', 'Richness', 'Sensitive richness', 
                    'Forage lichen abund', 'Cyanolichen abund', 
                    'Comm comp'), 
        caption= 'CL exceedances for mean nitrogen deposition (kg N ha^-1^ y^-1^), based on climate-adjusted lichen airscores.')
```

## Sulfur

### S exceedance: raw lichen airscores

```{r table_exceedance_s_A}
get_tab(pick = rx(), 
        xvar=c('tenyr', 's_raw_ex_sr.mean', 's_raw_ex_sr_sens.mean',
               's_raw_ex_sr_forag.mean', 's_raw_ex_sr_cyano.mean', 
               's_raw_ex_compn.mean'), 
        col.names=c('Ten-year interval', 'Richness', 'Sensitive richness', 
                    'Forage lichen abund', 'Cyanolichen abund', 
                    'Comm comp'), 
        caption= 'CL exceedances for mean sulfur deposition (kg S ha^-1^ y^-1^), based on raw lichen airscores.')
```

### S exceedance: climate-adjusted lichen airscores

```{r table_exceedance_s_B}
get_tab(pick = rx(), 
        xvar=c('tenyr', 's_adj_ex_sr.mean', 's_adj_ex_sr_sens.mean',   
               's_adj_ex_sr_forag.mean','s_adj_ex_sr_cyano.mean', 
               's_adj_ex_compn.mean'),
        col.names=c('Ten-year interval', 'Richness', 'Sensitive richness', 
                    'Forage lichen abund', 'Cyanolichen abund', 
                    'Comm comp'), 
        caption= 'CL exceedances for mean sulfur deposition (kg S ha^-1^ y^-1^), based on climate-adjusted lichen airscores.')
```


\newpage


# Further information

## Motivation

Each Wilderness area managed by the USDA Forest Service must report annually on progress toward “wildernesses meeting baseline performance for preserving wilderness character”, otherwise known as Wilderness Stewardship Performance (WSP).  Among other elements, **_Wilderness Air Quality Values_** (**_WAQVs_**) are a marker of natural quality of Wilderness character (WSP Guidebook 2020).  This report streamlines annual WSP reporting for WAQVs from lichen-based Critical Loads (CLs).

## Sensitive receptors

WAQVs are supported by monitoring changes in “sensitive receptors” like epiphytic macrolichens.  Lichens are impacted by air pollutants and therefore indicate changing conditions.  Workers long ago established links between lichen tissue content and the atmospheric deposition of pollutants like oxidized nitrogen (NO$_x$), reduced nitrogen (NH$_x$) and oxidized sulfur (SO$_x$).  Each species also has a unique and individualistic pollution tolerance, so observing collective responses is far more robust than any single species alone.

## Sensitive receptor indicators

For WSP reporting of air quality, “sensitive receptor indicators” can be estimated by the presence and abundance of lichen species.  Nationally-consistent FHM/FIA surveys of lichen communities let us score and classify species by their relative pollution tolerance, and to calculate community metrics across all species at a site.  Community metrics are more robust than tracking single species because the absence of a single species offers only circumstantial evidence of air quality (what if the species simply wasn't able to disperse to or tolerate microclimate at that site?).  The joint absence of multiple pollution-intolerant species provides much stronger evidence of air-quality impacts.

### Indicator: Air scores

Individual species' sensitivities are the value of N or S deposition (kg ha^-1^ y^-1^) at which the species was most frequently detected.  Air scores for each site are the community-weighted mean of species' sensitivities, for all species present (kg ha^-1^ y^-1^).  Therefore, site air scores represent the collective pollution tolerance of co-occurring species: "clean" sites have fairly low air scores driven by pollution-intolerant species, while "polluted" sites have higher air scores driven by pollution-tolerant ones.  Air scores are reported as both raw scores (without regard to climate) and as climate-adjusted scores (which permit nationwide comparison).

### Indicator: Critical loads exceedances

A critical load (CL) is the air pollutant deposition rate expected to result in harmful ecosystem changes.  Exceedance is the degree to which locally observed deposition goes beyond an established CL.  Lichen-based CLs were established in prior work (Geiser et al. 2019).  Specifically, lichen-based CLs are the deposition rates at which we expect changes in 1) total species richness, 2) sensitive species richness, 3) forage lichen abundance, 4) cyanolichen abundance, and 5) community compositions (see **[critical loads we used](#critical-loads-we-used)**).

## Conservation status

Conservation rankings follow NatureServe (https://www.natureserve.org/).  Species considered 'at risk' here are those ranked by NatureServe as Critically Imperiled, Imperiled, or Vulnerable, and/or those ranked by COSEWIC as Endangered, Special Concern, or Threatened.  

## Critical loads we used

We used critical loads from Table 3 in Geiser et al. (2019).  These critical loads are invariant to geography and so apply nationwide. Critical loads in this table are deposition levels resulting in $\geq20\%$ decrease in the richness or abundance of the given lichen metric. 

\  

```{r critical_loads_tab}
tab <- data.frame(Metric = c('Species richness',
                             'Sensitive species richness',
                             'Forage lichen abundance',
                             'Cyanolichen abundance',
                             'Comm composition shifts'),
                  Ndep = c(3.5, 3.1, 1.9, 1.3, 1.5),
                  Sdep = c(6.0, 2.5, 2.6, 2.3, 2.7))
knitr::kable(tab, format='pandoc', row.names = F, longtable = T,
             booktabs  = T,
             caption   = 'Critical loads used in this report (kg ha^-1^ y^-1^).',
             escape    = F,
             col.names = c('Lichen metric', 'N deposition', 'S deposition'))
```

See the article:  

>Geiser, L., P. Nelson, S. Jovan, H. Root, and C. Clark. 2019. Assessing ecological risks from atmospheric deposition of nitrogen and sulfur to US forests using epiphytic macrolichens. _Diversity_ 11 (87):1-29. [doi:10.3390/d11060087](https://doi.org/10.3390/d11060087){target="_blank"}  

\  

## Why monitor lichens? Ecological functions

\begin{wrapfigure}{r}{5.0cm}
\includegraphics{bothlichens.png}
\end{wrapfigure}

Lichens are commonly divided into "functional groups" based on ecological functions: three common groups are forage lichens, cyanolichens and matrix lichens.

__*Forage lichens*__ (including "old-man's beard" and "witches' hair" lichens) provide food, nesting materials and suitable habitat for many animals (from insects to amphibians to small mammals).   Large ungulates like deer, elk or caribou may depend on forage lichens to eat during lean winter months.  Loss of these pollution-sensitive lichens might have implications for the wildlife that depend on them.

__*Cyanolichens*__ (including "lungwort" lichens) contain cyanobacterial partners that convert nitrogen from the air into a form which forest vegetation can absorb -- when cyanolichens fall to the forest floor and release this nitrogen, they serve as "forest fertilizer"!  Cyanolichens are among the most intolerant of poor air quality, so their absence may indicate deteriorating conditions.

__*Matrix lichens*__ are any not included as forage or cyanolichens.  They are often more pollution-tolerant than those two groups, and perform an array of ecosystem functions.  These functions include rainfall interception, water cycle buffering, animal habitat, and bioindication of forest health.

Paying attention to changes in lichen communities is informative -- losses of lichen functional groups can indicate declines in the ecosystem functions that they provide.


\newpage


# Lichen species observed

## Sampling completeness

```{r run-species-completeness}
`get_richness` <- function (pick, ...) {
  NROW(s[[which(names(s) == pick)]])
}
spp_obs <- get_richness(pick = rx())
spp_exp <- 4.01 + 25.81*log(n) # log-linear eqn
```

Number of surveyed plots:   **`r n`**  
Number of species observed: **`r spp_obs`**  
Number of species expected: **`r round(spp_exp, 1)`** 

\  

```{r plot-sampling-completeness, fig.height=2.75, fig.cap='Species accumulation curve.  Expected value is based on a species accumulation curve (the line) where the expected number of species $y$ depends on sampling $x$ new plots: $y = 4.01 + 25.81 \\times log(x)$.'}
x <- 1:2500                              # n plots sampled
y <- 4.01 + 25.81*log(x)                 # n species expected
set_par(1)
plot(x, y, type='l', log='x', xaxt='n', xlim=c(1,2500),
     xlab=bquote(Plots~sampled~(log[10]~scale)),
     ylab='Species expected')
atx  <- outer(1:9, 10^(0:3))
labx <- ifelse(log10(atx) %% 1 == 0, atx, NA)
axis(1, at=atx, labels=labx, las=1)
lines(rep(n,2), c(spp_exp,spp_obs))
points(n, spp_exp, pch=21, cex=1.2, bg='white')
points(n, spp_obs, pch=16, cex=1.2)
legend('topleft', pch=c(16,21), bty='n', cex=0.7, pt.cex=1.0,
       legend=c(paste0('Observed = ', spp_obs),
                paste0('Expected = ', round(spp_exp,1))))
```

## Rare and pollution-sensitive species

```{r species-rare-sensitive}
`get_natureserve` <- function (pick, ...) {
  tab <- s[[which(names(s) == pick)]]
  rich   <- NROW(tab)
  nsens  <- round(sum(tab$n_class == 'oligotroph') / 
                    sum(tab$n_class != '') * 100, 1)
  ssens  <- round(sum(tab$s_class == 'sensitive')  / 
                    sum(tab$s_class != '') * 100, 1)
  atrisk <- sum(1 * grepl('1|2|3|Endanger|Concern|Threaten', 
                          tab$consrv_stat))
  list(rich=rich, nsens=nsens, ssens=ssens, atrisk=atrisk)
}
x <- get_natureserve(pick = rx())
```

Number of species observed in total: **`r x$rich`**  
Number of species with 'at risk' conservation status: **`r x$atrisk`**  
Percentage of rated species with 'sensitive (oligotroph)' N tolerance: **`r x$nsens`\%**  
Percentage of rated species with 'sensitive' S tolerance: **`r x$ssens`\%**  

## Species list

**List of all macrolichens in plots at `r input$x` from `r rng[1]` to `r rng[2]`**.  True diversity is likely greater because plot footprints are just a fraction of total Wilderness area.  _Functional groups_ ('Fxl grp') are cyanolichens ('cyano') and forage lichens ('forage')(see **[descriptions](#why-monitor-lichens-ecological-functions)**).  _Sensitivity_ for each species ranges from low to high tolerance for nitrogen (oligo-, meso-, eutroph) and sulfur (sensitive, intermediate, tolerant).  _Critical loads_ ('CL') are deposition levels resulting in >20\% decrease in occurrence.  _Conservation_ from NatureServe (https://www.natureserve.org/) assigns each species both a global (G) and state (S) rank on a scale of 1 to 5 (1=Critically imperiled; 2=Imperiled; 3=Vulnerable; 4=Apparently secure; 5=Secure).


```{r species_list}
`get_tab_species` <- function (pick, ...) {
  tab <- s[which(names(s) == pick)][[1]]
  tab <- tab[order(tab$n_critload, tab$s_critload, tab$fxlgp),]
  knitr::kable(x         = tab,
               format    = 'markdown',  ### format = 'pandoc',
               row.names = FALSE,
               longtable = TRUE,
               booktabs  = TRUE,
               caption   = NULL,  ### 'Macrolichen species list.',
               escape    = FALSE,
               col.names = c('Species','Fxl grp','N sensitivity','N CL',
                             'S sensitivity','S CL','Conservation')
  )
}
get_tab_species(pick = rx())
```

<!-- END -->