---
title: \vspace{-1.5cm} Wilderness Air Quality Values \break for `r input$x`
date: 'Report prepared: `r format(Sys.time(), "%d %B %Y")`'
header-includes:
- \usepackage{float} \floatplacement{figure}{H}
- \usepackage[font=small,skip=0pt]{caption}
- \usepackage{wrapfig}
- \usepackage{booktabs}
- \usepackage{longtable}
---

```{r setup-options, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '')
```

```{r function-setup-functions, include=FALSE}
### set graphical parameters
`set_par` <- function (panels = 1, pty = 's', ...) {
    mgp <- c(2, 0.4, 0)
    mar <- c(4, 4, 0.5, 0.5)
    oma <- c(0, 0, 0, 0)
    auto_rowcol <- function(n = panels) {
        if (n <= 3) 
            c(1, n)
        else if (n <= 6) 
            c(2, (n + 1)%/%2)
        else if (n <= 12) 
            c(3, (n + 2)%/%3)
        else c(ceiling(n/(nr <- ceiling(sqrt(n)))), nr)
    }
    mfrow <- auto_rowcol()
    if (panels > 4) 
        panels <- 4
    switch(as.character(panels), `4` = par(mfrow = mfrow, 
        mgp = mgp, mar = mar, pty = pty, oma = oma, bty = 'L', 
        las = 1, cex.lab = 1.2, tcl = -0.2, ...), `3` = par(mfrow = mfrow, 
        mgp = mgp, mar = mar, pty = pty, oma = oma, bty = 'L', 
        las = 1, cex.lab = 1.4, cex.axis = 1.2, tcl = -0.2, ...), 
        `2` = par(mfrow = mfrow, mgp = mgp, mar = mar, 
            pty = pty, oma = oma, bty = 'L', las = 1, cex.axis = 0.85, 
            tcl = -0.2, ...), `1` = par(mfrow = mfrow, 
            mgp = mgp, mar = mar, pty = pty, oma = oma, bty = 'L', 
            las = 1, cex.axis = 0.85, tcl = -0.2, ...))
}
### get years of sampling
`get_years` <- function (pick, ...) {
        rng <- range(d[d$wa==pick, 'year'], na.rm=TRUE)  ### <<--------------------
        if (any(!is.finite(rng))) c(NA,NA) else rng
}
### get recent decade
`get_recent_decade` <- function (pick, ...) {
     s <- sort(unique(a[a$warea==pick, 'tenyr']))
     s <- s[length(s)]
     if (any(!is.finite(s))) c(NA) else s
}
### get number of plots
`get_n` <- function (pick, ...) {
        n <- unique(a[a$warea==pick, 'n_plots'])
        if (any(!is.finite(n))) c(NA) else n
}
### table function with 5-y mean
`get_tab` <- function (pick, xvar, col.names, caption, format){
        tab <- a[which(a$warea==pick), xvar]
        if (missing(col.names)) col.names <- xvar
        if (missing(caption))   caption   <- NULL
        if (missing(format))    format    <- 'pandoc'
        j   <- xvar[xvar != 'tenyr'] # dont consider year column
        tab <- tab[rowSums(is.na(tab[,j])) != ncol(tab[,j]),]
        tab[tab %in% c('NA','***NA***','NaN','***NaN***')] <- NA 
        knitr::kable(x         = tab,
                     format    = format,
                     row.names = F,
                     align = c('l', rep('r',NCOL(tab)-1)),
                     longtable = T,
                     booktabs  = T,
                     caption   = caption,
                     col.names = col.names)
}
```

\vspace*{\fill}
\begin{wrapfigure}{l}{1.8cm}
\includegraphics{usfs.png}
\end{wrapfigure}

\  

Air Resource Management Program  
USDA Forest Service  
1400 Independence Ave SW  
Washington, DC 20250  

\  

\newpage

# Summary for `r input$x`

```{r years_and_plot_count}
rng <- get_years(pick = rx())
n   <- get_n(pick = rx())
dec <- get_recent_decade(pick = rx())
```

Number of surveyed plots:   **`r n`**  
Field sampling dates: **`r rng[1]` to `r rng[2]`**  
Most recent ten-year sampling period: **`r dec`**  
<!-- National rank for **critical loads exceedances**: *xx*^th^ percentile -->
<!-- National rank for **climate-adjusted lichen air scores**: *xx*^th^ percentile -->

## Nitrogen snapshot

```{r plot-plot_valuebar_N, fig.height=2.25, fig.width=6.0, fig.cap='Observed N deposition (arrow) for the most recent 10-year sample period, in relation to levels of concern.  Levels of concern (Acceptable, Marginal, Concerning, Poor, Very poor) are defined by the expected decline of sensitive lichens (0, 0-20, 20-50, 50-80, and >80 percent declines established for critical loads).'}
### plot the valuebar for observed CMAQ deposition
`plot_valuebar` <- function(pick, brk=b, select='n', ...){
     b      <- brk
     x      <- a[which(a$warea==pick),
                 paste0('cmaq_',select,'_3yroll.mean')]
     arrlab <- 'OBSERVED\nCMAQ deposition'
     if(is.na(x)) { # try lichen airscores if CMAQ is missing
          x      <- a[which(a$warea==pick),
                      paste0(select,'_airscore.mean')]
          arrlab <- 'OBSERVED\nlichen airscores'
     } 
     if(is.na(x)) { # no more tries
          x <- 999
     }
     obs <-  x[length(x)] # take most recent one
     par(mar=c(4,0.3,0.3,0.3), las=1, tcl=-0.2, mgp=c(1.8,0.4,0))
     n    <- 100
     xseq <- seq(-0.5, 20.0, len=n)
     plot(xseq, seq(1, 10, len=n), type='n', bty='n',
          yaxt='n', ylab='', ylim=c(1,13.4), yaxs='i', xaxs='r', ...)
     cc   <- viridis::inferno(5,begin=0.1,end=0.9,alpha=seq(0.7,1,len=5))
     rect(-9,   -10, b[1], 20, col=cc[1], border=NA)
     rect(b[1], -10, b[2], 20, col=cc[2], border=NA)
     rect(b[2], -10, b[3], 20, col=cc[3], border=NA)
     rect(b[3], -10, b[4], 20, col=cc[4], border=NA)
     rect(b[4], -10, 99,   20, col=cc[5], border=NA)
     text(x=b[1] - 0.5*(b[1]-(-1)), y=13.1, 'Acceptable', pos=1, cex=0.6)
     text(x=b[2] - 0.5*(b[2]-b[1]), y=13.1, 'Marginal', pos=1, cex=0.6)
     text(x=b[3] - 0.5*(b[3]-b[2]), y=13.1, 'Concerning', pos=1, cex=0.6)
     text(x=b[4] - 0.5*(b[4]-b[3]), y=13.1, 'Poor', pos=1, cex=0.6)
     text(x=21   - 0.5*(21-b[4]),   y=13.1, 'Very poor', pos=1, cex=0.6)
     if (obs >= 999) {
          text(x=10, y=10, 'No data available...', pos=1, cex=0.7, font=2)
     } else if (obs > 20 & obs < 999) {
          arrows(20.2, 6.5, 20.2, 1, length=0.20, angle=15, lwd=3, col=1)
          text(x=18.8, y=10, arrlab, pos=1, cex=0.7, font=2)
     } else {
          arrows(obs, 6.5, obs, 1, length=0.20, angle=15, lwd=3, col=1)
          text(x=obs, y=10, arrlab, pos=1, cex=0.7, font=2)
     }
}
b <- c(2.5, 6, 9.9, 14) # new (loosely based on 20/50/80)
plot_valuebar(pick = rx(), brk=b, select='n',
              xlab=expression(N~deposition~'('*kg~ha^-1~yr^-1*')'))
```

\  

## Sulfur snapshot

```{r plot_valuebar_S, fig.height=2.25, fig.width=6.0, fig.cap='Observed S deposition for the most recent 10-year sample period, in relation to levels of concern.  Levels of concern (Acceptable, Marginal, Concerning, Poor, Very poor) are defined by the expected decline of sensitive lichens (0, 0-20, 20-50, 50-80, and >80 percent declines established for critical loads).'}
### plot the valuebar for observed CMAQ S deposition
b <- b+(b*0.05) # new (loosely based on 20/50/80)
plot_valuebar(pick = rx(), brk=b, select='s',
              xlab=expression(S~deposition~'('*kg~ha^-1~yr^-1*')'))
```


\newpage


## Area map

```{r plot-map, fig.height=6.5, fig.width=6.49, fig.cap='Local area map.  Red outline is Wilderness boundary, black points are survey plot locations, shaded contours are terrain elevations above sea level.', warning=FALSE}
###  map wilderness boundary and sampling locations
`map_points` <- function (pick=pick, pts=d, shp=w, ras=dem, lwd.cont=1,
                          ...) {
     aea_prj <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5
        +lon_0=-110 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m"
     trg_prj <- projection(shp)
     shp  <- shp[shp$NAME == pick,]  # select your wilderness
     pts  <- pts[pts$wa == pick,]    # select your Wilderness
     if(all(is.na(pts$lon)) | all(is.na(pts$lon))) {
          # stop('Need lat/lon data to map')
          e <- extent(shp) * 1.25
          has_points <- FALSE
     } else {
          has_points <- TRUE
          `make_xy` <- function(xy, crs, ...){
               xy <- as.data.frame(xy)
               xy <- xy[!(is.na(xy$lon) | is.na(xy$lat)),] # rm NAs
               coordinates(xy) <- ~lon+lat
               proj4string(xy) <- CRS('+init=epsg:4269') # NAD83 dec degree
               return(spTransform(xy, crs))
          }
          ppts <- make_xy(pts, crs=trg_prj)
          near <- c(rgeos::gDistance(spTransform(ppts, CRS(aea_prj)),
                                     spTransform(shp, CRS(aea_prj)),
                                     byid=T) < 3000) # reject pts > 3 km
          ppts <- ppts[near,]
          e <- cbind(matrix(extent(shp)),matrix(extent(ppts)))
          e <- c(min(e[1,]),max(e[2,]),min(e[3,]),max(e[4,])) # combined ext
          `haversine` <- function(loc1, loc2) { # geographic distance
               lat1 <- loc1[1] * pi / 180
               lon1 <- loc1[2] * pi / 180
               lat2 <- loc2[1] * pi / 180
               lon2 <- loc2[2] * pi / 180
               R    <- 6371.0  # avg earth radius [km]
               dlat <- lat2 - lat1
               dlon <- lon2 - lon1
               a <- sin(dlat/2)^2 + cos(lat1) * cos(lat2) * sin(dlon/2)^2
               c <- 2 * atan2(sqrt(a), sqrt(1-a))
               return(R * c)
          }
          dy <- haversine(c(e[3],e[1]), c(e[4],e[1])) # kilometers y-axis
          dx <- haversine(c(e[3],e[1]), c(e[3],e[2])) # kilometers x-axis
          e  <- extent(e)
          if (dy > dx) {
               e[1:2] <- matrix(e * dy/dx)[1:2,] # expand x
          } else {
               e[3:4] <- matrix(e * dx/dy)[3:4,] # expand y
          }
          dx <- as.numeric( # how large a scale bar to make?
               as.character(
                    cut(dx,breaks=c(0,5,50,200,Inf),labels=c(1,5,10,50),incl=T)
               )
          )
          e  <- extent(e) * 1.25 # slight expansion
     }
     ras <- disaggregate(crop(ras, e), 10, method='bilinear')
     ### custom function for filled contours....
     `fillcont` <- function (x, maxpixels = 1e+05, ...) {
          `fc` <-  function (
               x = seq(0, 1, length.out = nrow(z)),
               y = seq(0, 1, length.out = ncol(z)),
               z,
               xlim = range(x, finite = TRUE),
               ylim = range(y, finite = TRUE),
               zlim = range(z, finite = TRUE),
               levels = pretty(zlim, nlevels),
               nlevels = 20,
               color.palette = function(n) viridis::viridis(n,begin=0,end=0.9),
               col = color.palette(length(levels) - 1), ...) {
               if (missing(z)) {
                    if (!missing(x)) {
                         if (is.list(x)) {
                              z <- x$z
                              y <- x$y
                              x <- x$x
                         }
                         else {
                              z <- x
                              x <- seq.int(0, 1, length.out = nrow(z))
                         }
                    }
                    else stop("no 'z' matrix specified")
               }
               else if (is.list(x)) {
                    y <- x$y
                    x <- x$x
               }
               if (any(diff(x) <= 0) || any(diff(y) <= 0))
                    stop("increasing 'x' and 'y' values expected")
               plot.new()
               plot.window(xlim, ylim, "", xaxs = "i", yaxs = "i", asp = NA)
               .filled.contour(x, y, z, levels, col)
          }
          x <- sampleRegular(x, maxpixels, asRaster = T, useGDAL = T)
          X <- xFromCol(x, 1:ncol(x))
          Y <- yFromRow(x, nrow(x):1)
          Z <- t(matrix(getValues(x), ncol = x@ncols, byrow=T)[nrow(x):1,])
          lonlat <- couldBeLonLat(x, warnings = FALSE)
          asp <- list(...)$asp
          if (is.null(asp)) {
               if (lonlat) {
                    ym <- mean(c(x@extent@ymax, x@extent@ymin))
                    asp <- 1/cos((ym * pi)/180)
               }
               else {
                    asp <- 1
               }
               fc(x = X, y = Y, z = Z, asp = asp, ...)
          }
          else {
               fc(x = X, y = Y, z = Z, ...)
          }
     }
     fillcont(ras, nlevels=20, lwd=lwd.cont)
     plot(shp, col='transparent', border='red', add=T, lwd=2)
     if (has_points) {
          points(ppts, pch=16, col='#000000', cex=1)
          scalebar(d=dx, xy=c(e[1] + abs(diff(e[1:2]))*0.05,
                              e[3] + abs(diff(e[3:4]))*0.08),
                   type='bar', divs=2, below=paste0('\n', dx, ' km'),
                   adj=c(0.5,-1), lab='', lwd=2, cex=0.7, font=2)    
     }
}
# `map_points` <- function (pick=pick, pts=d, shp=w, ras=dem, lwd.cont=1,
#                           ...) {
#     aea_prj <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5
#         +lon_0=-110 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m"
#     trg_prj <- projection(shp)
#     shp  <- shp[shp$NAME == pick,]  # select your wilderness
#     pts  <- pts[pts$wa == pick,]    # select your Wilderness
#     if(all(is.na(pts$lon)) | all(is.na(pts$lon))) {
#         stop('Need lat/lon data to map')
#     }
#     `make_xy` <- function(xy, crs, ...){
#         xy <- as.data.frame(xy)
#         xy <- xy[!(is.na(xy$lon) | is.na(xy$lat)),] # rm NAs
#         coordinates(xy) <- ~lon+lat
#         proj4string(xy) <- CRS('+init=epsg:4269') # NAD83 dec degree
#         return(spTransform(xy, crs))
#     }
#     ppts <- make_xy(pts, crs=trg_prj)
#     near <- c(rgeos::gDistance(spTransform(ppts, CRS(aea_prj)),
#                                spTransform(shp, CRS(aea_prj)),
#                                byid=T) < 3000) # reject pts > 3 km
#     ppts <- ppts[near,]
#     e <- cbind(matrix(extent(shp)),matrix(extent(ppts)))
#     e <- c(min(e[1,]),max(e[2,]),min(e[3,]),max(e[4,])) # combined ext
#     `haversine` <- function(loc1, loc2) { # geographic distance
#         lat1 <- loc1[1] * pi / 180
#         lon1 <- loc1[2] * pi / 180
#         lat2 <- loc2[1] * pi / 180
#         lon2 <- loc2[2] * pi / 180
#         R    <- 6371.0  # avg earth radius [km]
#         dlat <- lat2 - lat1
#         dlon <- lon2 - lon1
#         a <- sin(dlat/2)^2 + cos(lat1) * cos(lat2) * sin(dlon/2)^2
#         c <- 2 * atan2(sqrt(a), sqrt(1-a))
#         return(R * c)
#     }
#     dy <- haversine(c(e[3],e[1]), c(e[4],e[1])) # kilometers y-axis
#     dx <- haversine(c(e[3],e[1]), c(e[3],e[2])) # kilometers x-axis
#     e  <- extent(e)
#     if (dy > dx) {
#         e[1:2] <- matrix(e * dy/dx)[1:2,] # expand x
#     } else {
#         e[3:4] <- matrix(e * dx/dy)[3:4,] # expand y
#     }
#     dx <- as.numeric( # how large a scale bar to make?
#         as.character(
#             cut(dx,breaks=c(0,5,50,200,Inf),labels=c(1,5,10,50),incl=T)
#         )
#     )
#     e  <- extent(e) * 1.25 # slight expansion
#     ras <- disaggregate(crop(ras, e), 10, method='bilinear')
#     ### custom function for filled contours....
#     `fillcont` <- function (x, maxpixels = 1e+05, ...) {
#         `fc` <-  function (
#             x = seq(0, 1, length.out = nrow(z)),
#             y = seq(0, 1, length.out = ncol(z)),
#             z,
#             xlim = range(x, finite = TRUE),
#             ylim = range(y, finite = TRUE),
#             zlim = range(z, finite = TRUE),
#             levels = pretty(zlim, nlevels),
#             nlevels = 20,
#             color.palette = function(n) viridis::viridis(n, begin=0, end=0.9),
#             col = color.palette(length(levels) - 1), ...) {
#             if (missing(z)) {
#                 if (!missing(x)) {
#                     if (is.list(x)) {
#                         z <- x$z
#                         y <- x$y
#                         x <- x$x
#                     }
#                     else {
#                         z <- x
#                         x <- seq.int(0, 1, length.out = nrow(z))
#                     }
#                 }
#                 else stop("no 'z' matrix specified")
#             }
#             else if (is.list(x)) {
#                 y <- x$y
#                 x <- x$x
#             }
#             if (any(diff(x) <= 0) || any(diff(y) <= 0))
#                 stop("increasing 'x' and 'y' values expected")
#             plot.new()
#             plot.window(xlim, ylim, "", xaxs = "i", yaxs = "i", asp = NA)
#             .filled.contour(x, y, z, levels, col)
#         }
#         x <- sampleRegular(x, maxpixels, asRaster = T, useGDAL = T)
#         X <- xFromCol(x, 1:ncol(x))
#         Y <- yFromRow(x, nrow(x):1)
#         Z <- t(matrix(getValues(x), ncol = x@ncols, byrow=T)[nrow(x):1,])
#         lonlat <- couldBeLonLat(x, warnings = FALSE)
#         asp <- list(...)$asp
#         if (is.null(asp)) {
#             if (lonlat) {
#                 ym <- mean(c(x@extent@ymax, x@extent@ymin))
#                 asp <- 1/cos((ym * pi)/180)
#             }
#             else {
#                 asp <- 1
#             }
#             fc(x = X, y = Y, z = Z, asp = asp, ...)
#         }
#         else {
#             fc(x = X, y = Y, z = Z, ...)
#         }
#     }
#     # contour(ras, col=viridis::viridis(11,1,0,0.9), nlevels=10,
#     #         drawlabels=F, axes=F, lwd=lwd.cont)
#     fillcont(ras, nlevels=20, lwd=lwd.cont)
#     plot(shp, col='transparent', border='red', add=T, lwd=2)
#     points(ppts, pch=16, col='#000000', cex=1)
#     scalebar(d=dx, xy=c(e[1] + abs(diff(e[1:2]))*0.05,
#                         e[3] + abs(diff(e[3:4]))*0.08),
#              type='bar', divs=2, below=paste0('\n', dx, ' km'),
#              adj=c(0.5,-1), lab='', lwd=2, cex=0.7, font=2)
# }
map_points(pick = rx(), pts=d, shp=w, ras=dem)
```


\newpage


# Site scores

You can think of site scores (or "air scores") as the "average" position of species along an air quality gradient.  Technically, these are the community-weighted mean of species' sensitivities, where sensitivities are defined as the value of N or S deposition (kg ha^-1^ y^-1^) at which each species was most frequently detected.  We report both raw and climate-adjusted scores, and their nationwide ranking as a percentile.^[See **["Indicator: Air scores"](#indicator-air-scores)** section for more information.].

## Nitrogen

### N: mean deposition
```{r table_n_site}
get_tab(pick = rx(), 
        xvar=c('tenyr','n_airscore.mean','nadj.mean','cmaq_n_3yroll.mean'), 
        col.names=c('Ten-year interval', 'Lichen airscore (raw)', 'Lichen airscore (adj)', 'CMAQ model'), 
        caption= 'Site scores for estimated nitrogen deposition (N), as a *mean* (kg ha^-1^ y^-1^).')
```

### N: percentiles
```{r table_n_site_perc}
get_tab(pick = rx(), 
        xvar=c('tenyr','p_n_airscore.mean','p_n_adj.mean','p_n_cmaq.mean'), 
        col.names=c('Ten-year interval', 'Lichen airscore (raw)', 'Lichen airscore (adj)', 'CMAQ model'),
        caption= 'Site scores for estimated nitrogen deposition (N), as a *percentile* (0-100).')
```

## Sulfur

### S: mean deposition
```{r table_s_site}
get_tab(pick = rx(), 
        xvar=c('tenyr','s_airscore.mean','sadj.mean','cmaq_s_3yroll.mean'), 
        col.names=c('Ten-year interval', 'Lichen airscore (raw)', 'Lichen airscore (adj)', 'CMAQ model'), 
        caption= 'Site scores for estimated sulfur deposition (S), as a *mean* (kg ha^-1^ y^-1^).')
```

### S: percentiles
```{r table_s_site_perc}
get_tab(pick = rx(), 
        xvar=c('tenyr','p_s_airscore.mean','p_s_adj.mean','p_s_cmaq.mean'), 
        col.names=c('Ten-year interval', 'Lichen airscore (raw)', 'Lichen airscore (adj)', 'CMAQ model'),
        caption= 'Site scores for estimated sulfur deposition (S), as a *percentile* (0-100).')
```


\newpage


# Critical loads exceedances

Exceedances are the departure of raw or climate-adjusted lichen airscores from established lichen-based critical loads^[See **["Critical loads we used"](#critical-loads-we-used)** section.], expressed as the mean of a ten-year period.  Negative values indicate the "safety margin" below the critical load threshold, positive values indicate exceedance above the threshold. 

## Nitrogen

### N exceedance: raw lichen airscores

```{r table_exceedance_n_A}
get_tab(pick = rx(), 
        xvar=c('tenyr', 'n_raw_ex_sr.mean', 'n_raw_ex_sr_sens.mean',
               'n_raw_ex_sr_forag.mean', 'n_raw_ex_sr_cyano.mean', 
               'n_raw_ex_compn.mean'), 
        col.names=c('Ten-year interval', 'Richness', 'Sensitive richness', 
                    'Forage lichen abund', 'Cyanolichen abund', 
                    'Comm comp'), 
        caption= 'CL exceedances for mean nitrogen deposition (kg N ha^-1^ y^-1^), based on raw lichen airscores.')
```

### N exceedance: climate-adjusted lichen airscores

```{r table_exceedance_n_B}
get_tab(pick = rx(), 
        xvar=c('tenyr', 'n_adj_ex_sr.mean',       
               'n_adj_ex_sr_sens.mean',   
               'n_adj_ex_sr_forag.mean',  
               'n_adj_ex_sr_cyano.mean', 
               'n_adj_ex_compn.mean'),
        col.names=c('Ten-year interval', 'Richness', 'Sensitive richness', 
                    'Forage lichen abund', 'Cyanolichen abund', 
                    'Comm comp'), 
        caption= 'CL exceedances for mean nitrogen deposition (kg N ha^-1^ y^-1^), based on climate-adjusted lichen airscores.')
```

## Sulfur

### S exceedance: raw lichen airscores

```{r table_exceedance_s_A}
get_tab(pick = rx(), 
        xvar=c('tenyr', 's_raw_ex_sr.mean', 's_raw_ex_sr_sens.mean',
               's_raw_ex_sr_forag.mean', 's_raw_ex_sr_cyano.mean', 
               's_raw_ex_compn.mean'), 
        col.names=c('Ten-year interval', 'Richness', 'Sensitive richness', 
                    'Forage lichen abund', 'Cyanolichen abund', 
                    'Comm comp'), 
        caption= 'CL exceedances for mean sulfur deposition (kg S ha^-1^ y^-1^), based on raw lichen airscores.')
```

### S exceedance: climate-adjusted lichen airscores

```{r table_exceedance_s_B}
get_tab(pick = rx(), 
        xvar=c('tenyr', 's_adj_ex_sr.mean', 's_adj_ex_sr_sens.mean',   
               's_adj_ex_sr_forag.mean','s_adj_ex_sr_cyano.mean', 
               's_adj_ex_compn.mean'),
        col.names=c('Ten-year interval', 'Richness', 'Sensitive richness', 
                    'Forage lichen abund', 'Cyanolichen abund', 
                    'Comm comp'), 
        caption= 'CL exceedances for mean sulfur deposition (kg S ha^-1^ y^-1^), based on climate-adjusted lichen airscores.')
```


\newpage


# Further information

## Motivation

Each Wilderness area managed by the USDA Forest Service must report annually on progress toward “wildernesses meeting baseline performance for preserving wilderness character”, otherwise known as Wilderness Stewardship Performance (WSP).  Among other elements, **_Wilderness Air Quality Values_** (**_WAQVs_**) are a marker of natural quality of Wilderness character (WSP Guidebook 2020).  This report streamlines annual WSP reporting for WAQVs from lichen-based Critical Loads (CLs).
<!-- This report enables nationally consistent, accurate, and timely annual reporting for WAQVs from lichen-based Critical Loads (CLs). -->

## Sensitive receptors

WAQVs are supported by monitoring changes in “sensitive receptors” like epiphytic macrolichens.  Lichens are impacted by air pollutants and therefore indicate changing conditions.  Workers long ago established links between lichen tissue content and the atmospheric deposition of pollutants like oxidized nitrogen (NO$_x$), reduced nitrogen (NH$_x$) and oxidized sulfur (SO$_x$).  Each species also has a unique and individualistic pollution tolerance, so observing collective responses is far more robust than any single species alone.

## Sensitive receptor indicators

For WSP reporting of air quality, “sensitive receptor indicators” can be estimated by the presence and abundance of lichen species.  Nationally-consistent FHM/FIA surveys of lichen communities let us score and classify species by their relative pollution tolerance, and to calculate community metrics across all species at a site.  Community metrics are more robust than tracking single species because the absence of a single species offers only circumstantial evidence of air quality (what if the species simply wasn't able to disperse to or tolerate microclimate at that site?).  The joint absence of multiple pollution-intolerant species provides much stronger evidence of air-quality impacts.

### Indicator: Air scores

Individual species' sensitivities are the value of N or S deposition (kg ha^-1^ y^-1^) at which the species was most frequently detected.  Air scores for each site are the community-weighted mean of species' sensitivities, for all species present (kg ha^-1^ y^-1^).  Therefore, site air scores represent the collective pollution tolerance of co-occurring species: "clean" sites have fairly low air scores driven by pollution-intolerant species, while "polluted" sites have higher air scores driven by pollution-tolerant ones.  Air scores are reported as both raw scores (without regard to climate) and as climate-adjusted scores (which permit nationwide comparison).

### Indicator: Critical loads exceedances

A critical load (CL) is the air pollutant deposition rate expected to result in harmful ecosystem changes.  Exceedance is the degree to which locally observed deposition goes beyond an established CL.  Lichen-based CLs were established in prior work (Geiser et al. 2019).  Specifically, lichen-based CLs are the deposition rates at which we expect changes in 1) total species richness, 2) sensitive species richness, 3) forage lichen abundance, 4) cyanolichen abundance, and 5) community compositions (see **[next section](#critical-loads-we-used)**).

## Conservation status

Conservation rankings follow NatureServe (https://www.natureserve.org/).  Species considered 'at risk' here are those ranked by NatureServe as Critically Imperiled, Imperiled, or Vulnerable, and/or those ranked by COSEWIC as Endangered, Special Concern, or Threatened.  

## Critical loads we used

We used critical loads from Table 3 in Geiser et al. (2019).  These critical loads are invariant to geography and so apply nationwide. Critical loads in this table are deposition levels resulting in $\geq20\%$ decrease in the richness or abundance of the given lichen metric. 


\  


```{r critical_loads_tab}
tab <- data.frame(Metric = c('Species richness',
                             'Sensitive species richness',
                             'Forage lichen abundance',
                             'Cyanolichen abundance',
                             'Comm composition shifts'),
                  Ndep = c(3.5, 3.1, 1.9, 1.3, 1.5),
                  Sdep = c(6.0, 2.5, 2.6, 2.3, 2.7))
knitr::kable(tab, format='pandoc', row.names = F, longtable = T,
             booktabs  = T,
             caption   = 'Critical loads used in this report (kg ha^-1^ y^-1^).',
             escape    = F,
             col.names = c('Lichen metric', 'N deposition', 'S deposition'))
```

See the article:  

>Geiser, L., P. Nelson, S. Jovan, H. Root, and C. Clark. 2019. Assessing ecological risks from atmospheric deposition of nitrogen and sulfur to US forests using epiphytic macrolichens. _Diversity_ 11 (87):1-29. [doi:10.3390/d11060087](https://doi.org/10.3390/d11060087){target="_blank"}  

<!-- ### Trophic scores -->
<!-- Lichen species can be assigned to three main groups based on pollution tolerance: oligo-, meso- and eutrophic (from low to high tolerance). Sites can be scored by the ratio of oligotrophic species richness _vs._ the total richness of oligotrophic and eutrophic species.  By this metric, relatively "clean" sites have high trophic scores (maximum = 1) driven by pollution-intolerant species, while "polluted" sites have lower trophic scores (minimum = 0) driven by their absence. Trophic scores are unitless and range from 0--1 for N or S. -->

\  

## Why monitor lichens? Ecological functions

Lichens are commonly divided into three "functional groups" based on ecological functions: matrix lichens, forage lichens, and cyanolichens.

Forage lichens (including "old-man's beard" and "witches' hair" lichens) provide food, nesting materials and suitable habitat for many animals (from insects to amphibians to small mammals).   Large ungulates like deer, elk or caribou may depend on forage lichens to eat during lean winter months.  Loss of these pollution-sensitive lichens might have implications for the wildlife that depend on them.

Cyanolichens (including "lungwort" lichens) contain cyanobacterial partners that convert nitrogen from the air into a form which forest vegetation can absorb -- when cyanolichens fall to the forest floor and release this nitrogen, they serve as "forest fertilizer"! Despite this capacity, cyanolichens are among the most intolerant of poor air quality, and so their presence indicates clean air.

Matrix lichens are any that are not forage or cyanolichens.  They are often more pollution-tolerant than those two groups, and perform an array of ecosystem functions. These range from...


\newpage


# Lichen species observed

## Sampling completeness

```{r run-species-completeness}
`get_richness` <- function (pick, ...) {
  NROW(s[[which(names(s) == pick)]])
}
# `get_n_plots` <- function (pick, ...) {
#   NROW(d[which(d$wa==pick),])
# }
spp_obs    <- get_richness(pick = rx())
# plot_obs   <- n
spp_exp    <- 4.01 + 25.81*log(n) # log-linear eqn
```

Number of surveyed plots:   **`r n`**  
Number of species observed: **`r spp_obs`**  
Number of species expected: **`r round(spp_exp, 1)`** 

\  

```{r plot-sampling-completeness, fig.height=2.75, fig.cap='Species accumulation curve.  Expected value is based on a species accumulation curve (the line) where the expected number of species $y$ depends on sampling $x$ new plots: $y = 4.01 + 25.81 \\times log(x)$.'}
x <- 1:2500                              # n plots sampled
y <- 4.01 + 25.81*log(x)                 # n species expected
set_par(1)
plot(x, y, type='l', log='x', xaxt='n', xlim=c(1,2500),
     xlab=bquote(Plots~sampled~(log[10]~scale)),
     ylab='Species expected')
atx  <- outer(1:9, 10^(0:3))
labx <- ifelse(log10(atx) %% 1 == 0, atx, NA)
axis(1, at=atx, labels=labx, las=1)
lines(rep(n,2), c(spp_exp,spp_obs))
points(n, spp_exp, pch=21, cex=1.2, bg='white')
points(n, spp_obs, pch=16, cex=1.2)
legend('topleft', pch=c(16,21), bty='n', cex=0.7, pt.cex=1.0,
       legend=c(paste0('Observed = ', spp_obs),
                paste0('Expected = ', round(spp_exp,1))))
```

## Rare and pollution-sensitive species

```{r species-rare-sensitive}
`get_natureserve` <- function (pick, ...) {
  tab <- s[[which(names(s) == pick)]]
  rich   <- NROW(tab)
  nsens  <- round(sum(tab$n_class == 'oligotroph') / 
                    sum(tab$n_class != '') * 100, 1)
  ssens  <- round(sum(tab$s_class == 'sensitive')  / 
                    sum(tab$s_class != '') * 100, 1)
  atrisk <- sum(1 * grepl('1|2|3|Endanger|Concern|Threaten', 
                          tab$consrv_stat))
  list(rich=rich, nsens=nsens, ssens=ssens, atrisk=atrisk)
}
x <- get_natureserve(pick = rx())
```

Number of species observed in total: **`r x$rich`**  
Number of species with 'at risk' conservation status: **`r x$atrisk`**  
Percentage of rated species with 'sensitive (oligotroph)' N tolerance: **`r x$nsens`\%**  
Percentage of rated species with 'sensitive' S tolerance: **`r x$ssens`\%**  

## Species list

**List of all macrolichens observed in plots at `r input$x` over the years `r rng[1]` to `r rng[2]`**.  True diversity is likely greater because plot footprints are just a fraction of total Wilderness area.  _Functional groups_ ('Fxl grp'): include cyanolichens ('cyano') capable of nitrogen-fixation, and forage lichens ('forage') important for wildlife.  _Sensitivity_: 'N sensitivity' for each species ranges from low to high tolerance (oligo-, meso-, eutroph), as does 'S sensitivity' (sensitive, intermediate, tolerant).  _Critical loads_ ('CL') are deposition levels resulting in >20\% decrease in occurrence.  _Conservation_: conservation status is from NatureServe (https://www.natureserve.org/), which assigns each species both a global (G) and state (S) rank on a scale of 1 to 5 (1=Critically imperiled; 2=Imperiled; 3=Vulnerable; 4=Apparently secure; 5=Secure).  

```{r species_list}
`get_tab_species` <- function (pick, ...) {
  tab <- s[which(names(s) == pick)]
  knitr::kable(x         = tab,
               format    = 'markdown',  ### format = 'pandoc',
               row.names = F,
               longtable = T,
               booktabs  = T,
               caption   = 'Macrolichen species list.',
               escape    = F,
               col.names = c('Species','Fxl grp','N sensitivity','N CL',
                             'S sensitivity','S CL','Conservation')
  )
}
get_tab_species(pick = rx())
```

End of report.
<!-- END -->